---
title: "R Markdown for Scientists — Welcome & Setup"
subtitle: "NIEHS 3-Hour Workshop"
author: "NIEHS Training"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Welcome! {.unnumbered}

Welcome to **R Markdown for Scientists**, a hands-on workshop designed for
researchers at the National Institute of Environmental Health Sciences (NIEHS).

By the end of this 3-hour training, you will be able to:

- Write reproducible reports that combine prose, code, and results
- Control how code and output appear in your documents
- Create publication-quality tables and figures
- Add equations, citations, and cross-references
- Knit documents to HTML, Word, and PDF formats

## Workshop Agenda

| Module | Topic | Time |
|--------|-------|------|
| 0 | Welcome & Setup (this notebook) | 10 min |
| 1 | R Markdown Basics: YAML, Markdown, and Knitting | 25 min |
| 2 | Code Chunks and Inline Code | 30 min |
| 3 | Tables and Data Display | 25 min |
| — | Break | 10 min |
| 4 | Visualizations and Figure Control | 25 min |
| 5 | Scientific Writing: Equations, Citations, Parameters | 20 min |
| 6 | Output Formats, Themes, and Templates | 15 min |
| 7 | Capstone: Build a Complete Analysis Report | 20 min |

## What is R Markdown?

R Markdown is a file format (`.Rmd`) that lets you combine:
 
1. **Narrative text** written in Markdown (a simple formatting language)
2. **R code** that runs and produces results
3. **Output** — figures, tables, and statistics — embedded directly in the document

When you **knit** an `.Rmd` file, R executes all the code, inserts the results,
and produces a polished report in HTML, Word, PDF, or other formats.

### Why R Markdown for NIEHS Research?

- **Reproducibility**: Your report IS the analysis. Anyone can re-run it and get
  the same results.
- **FAIR principles**: Code, data, and narrative are bundled together in a
  transparent, shareable format.
- **Efficiency**: Change your data or methods, re-knit, and the entire report
  updates — no copy-paste errors.
- **Collaboration**: Share `.Rmd` files via GitHub and your whole team can see
  exactly how results were generated.

# Setting Up Your Environment

## Posit Cloud Prerequisites

You should already have:

- A **Posit Cloud** account (free tier is fine) — [posit.cloud](https://posit.cloud)
- A project open in Posit Cloud (create a new one if needed)

## Installing Required Packages

Run the code chunk below to install the packages we will use throughout the
workshop. **This may take a few minutes on Posit Cloud** — that is normal.

```{r install-packages, eval=FALSE}
# Run this once — you do NOT need to run it every time you open the project
install.packages(c(
  "rmarkdown",    # Core R Markdown engine (likely already installed)
  "knitr",        # Code chunk processing
  "tidyverse",    # Data manipulation and ggplot2
  "kableExtra",   # Beautiful tables
  "DT",           # Interactive HTML tables
  "plotly",       # Interactive plots
  "broom",        # Tidy model output
  "here"          # Portable file paths
))
```

> **Posit Cloud note**: The `rmarkdown` and `knitr` packages come pre-installed on
> Posit Cloud projects. You mainly need to install the table and visualization
> packages listed above.

## Verify Your Setup

Run this chunk to confirm everything is working:

```{r verify-setup}
# Check that key packages load
library(rmarkdown)
library(knitr)
library(tidyverse)

cat("R version:", R.version.string, "\n")
cat("rmarkdown version:", as.character(packageVersion("rmarkdown")), "\n")
cat("knitr version:", as.character(packageVersion("knitr")), "\n")
cat("\n✓ Setup looks good! You're ready to begin.\n")
```

# How to Use These Notebooks

Each notebook in this workshop is itself an R Markdown document. You are
learning R Markdown *by using R Markdown*.

**Two ways to work through the material:**

1. **Run chunks interactively**: Click the green "play" button (▶) on any code
   chunk to run it in place and see the results below the chunk.
2. **Knit the whole document**: Click the **Knit** button (or press
   `Ctrl+Shift+K` / `Cmd+Shift+K`) to render the entire notebook into an HTML
   report.

**Try it now!** Click the **Knit** button at the top of this document to see what
the rendered output looks like.

# Quick Tour of the RStudio Interface

When you have an `.Rmd` file open, notice these key areas:

- **Source pane** (upper left): Where you edit the `.Rmd` file
- **Knit button**: At the top of the source pane — click to render
- **Run buttons**: Green arrows on each code chunk to run interactively
- **Console** (lower left): Where R code executes when you run chunks
- **Output/Viewer** (lower right): Where knitted HTML appears

## A Note on PDF Output

Throughout this workshop we will primarily knit to **HTML**, which works
perfectly on Posit Cloud with no extra setup.

If you want to knit to **PDF**, you need a LaTeX installation. On Posit Cloud
you can install TinyTeX:

```{r install-tinytex, eval=FALSE}
# Only run this if you want PDF output (optional — takes several minutes)
install.packages("tinytex")
tinytex::install_tinytex()
```

> We recommend sticking with HTML output during the workshop. It is faster, 
> supports interactive features (like clickable tables), and avoids LaTeX
> troubleshooting.

---

**You're all set!** Open the next notebook: `01_rmarkdown_basics.Rmd`
