---
title: "Module 2: Code Chunks and Inline Code"
subtitle: "Controlling How Code Runs and What Readers See"
author: "NIEHS Training"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

# Code Chunks: The Heart of R Markdown

Code chunks are fenced blocks of R code that run when you knit the document.
They look like this in the raw `.Rmd` file:

````
```{r chunk-name}`r ''`
# Your R code goes here
mean(c(1, 2, 3, 4, 5))
```
````

Every chunk starts with `` ```{r} `` and ends with `` ``` ``.

## Creating Code Chunks

Three ways to insert a new code chunk:

1. **Keyboard shortcut**: `Ctrl+Alt+I` (Windows/Linux) or `Cmd+Option+I` (Mac)
2. **Menu**: Code → Insert Chunk
3. **Type it manually**: Type the opening and closing fences

## Naming Your Chunks

The text after `r` is the chunk name (optional but recommended):

````
```{r load-exposure-data}`r ''`
exposure <- read.csv("data/exposure.csv")
```
````

Good chunk names help you:

- Navigate quickly via the document outline (bottom-left of source pane)
- Identify errors when knitting fails ("Error in chunk 'load-exposure-data'")
- Cache results by chunk name

> **Rule:** Chunk names must be unique within a document. Duplicate names will
> cause a knitting error.

## Running Chunks

- **Run current chunk**: Click the green ▶ button, or `Ctrl+Shift+Enter`
- **Run all chunks above**: Click the downward arrow icon on the chunk toolbar
- **Run a single line**: Place cursor on the line and press `Ctrl+Enter`


# Chunk Options: Controlling Output

Chunk options let you control what appears in the final document. They go inside
the `{r}` header, separated by commas.

## The Most Important Chunk Options

### `echo` — Show or hide the code

```{r echo-true, echo=TRUE}
# echo=TRUE (default) — readers see the code AND the output
result <- 2 + 2
result
```

```{r echo-false, echo=FALSE}
# echo=FALSE — readers see ONLY the output, not the code
cat("This result appeared without showing the code that produced it.")
```

Use `echo=FALSE` for final reports where readers care about results, not code.

### `eval` — Run or skip the code

```{r eval-demo, eval=FALSE}
# eval=FALSE — the code is displayed but NOT executed
# Useful for showing example code that shouldn't actually run
install.packages("some_package")  # This won't actually run
```

### `include` — Show anything at all?

```{r include-demo, include=FALSE}
# include=FALSE — the code runs silently (no code, no output shown)
# Perfect for setup steps like loading packages
secret_value <- 42
```

The chunk above ran silently. Proof: the value of `secret_value` is
`r secret_value`.

### `message` and `warning` — Suppress messages

```{r with-message, message=TRUE}
library(tidyverse)  # This would normally show package startup messages
```

Set `message=FALSE` and `warning=FALSE` to keep your output clean. We already
did this globally in our setup chunk (scroll to the top of this file).

## Common Chunk Option Combinations

| Scenario | Options |
|----------|---------|
| Show code and output (default) | `echo=TRUE, eval=TRUE` |
| Show results only (for reports) | `echo=FALSE` |
| Show code but don't run it | `eval=FALSE` |
| Run silently (setup, data loading) | `include=FALSE` |
| Display a warning-free figure | `echo=FALSE, warning=FALSE` |

## Exercise 2.1: Practice Chunk Options {.unnumbered}

**Try this:** Modify the chunk below to hide the code but show the result.
Change `echo=TRUE` to `echo=FALSE`, then knit.

```{r exercise-echo, echo=TRUE}
# Environmental data summary
chemicals <- c("PFOA", "PFOS", "PFNA", "PFHxS", "GenX")
half_life_years <- c(3.8, 5.4, 2.5, 8.5, NA)

data.frame(
  Chemical = chemicals,
  `Half-life (years)` = half_life_years,
  check.names = FALSE
)
```


# Global Chunk Options

Instead of setting options on every chunk, you can set **defaults** for the
entire document using `knitr::opts_chunk$set()` in a setup chunk:

````
```{r setup, include=FALSE}`r ''`
knitr::opts_chunk$set(
  echo = TRUE,        # Show code by default
  message = FALSE,    # Suppress messages by default
  warning = FALSE,    # Suppress warnings by default
  fig.width = 7,      # Default figure width (inches)
  fig.height = 5,     # Default figure height (inches)
  fig.align = "center" # Center all figures
)
```
````

**This is standard practice.** Every R Markdown document you write should start
with a setup chunk. Look at the top of this file — we have one.

You can always override global options on individual chunks when needed.


# Inline Code

Inline code lets you embed R results directly in your text. This is incredibly
useful for reporting statistics.

## Syntax

Use a single backtick with `r`:

```
The mean exposure level was `r knitr::inline_expr('mean(c(2.3, 4.1, 3.7))')` µg/L.
```

Renders as:

The mean exposure level was `r mean(c(2.3, 4.1, 3.7))` µg/L.

## Why Inline Code Matters

Imagine you are writing a report and your data changes. Without inline code:

> "We observed 247 participants with a mean age of 43.2 years (SD = 12.1)."

You would need to find and update every number manually. With inline code, the
numbers **update automatically** when you re-knit.

## A Realistic Example

```{r inline-data}
# Simulated NIEHS study data
set.seed(42)
n_participants <- 150
ages <- round(rnorm(n_participants, mean = 45, sd = 12))
bmi <- round(rnorm(n_participants, mean = 27.3, sd = 4.5), 1)
pfoa_serum <- round(rlnorm(n_participants, meanlog = 1.2, sdlog = 0.8), 2)
```

Now we can write a paragraph using inline code:

> Our study enrolled `r n_participants` participants with a mean age of
> `r round(mean(ages), 1)` years (SD = `r round(sd(ages), 1)`). The median
> serum PFOA concentration was `r round(median(pfoa_serum), 2)` ng/mL
> (IQR: `r round(quantile(pfoa_serum, 0.25), 2)`–`r round(quantile(pfoa_serum, 0.75), 2)`).
> Mean BMI was `r round(mean(bmi), 1)` kg/m² (SD = `r round(sd(bmi), 1)`).

**Every number in that paragraph came from R code.** If we change the sample
size, re-knit, and all the statistics update.

## Formatting Inline Numbers

For cleaner output, round and format your numbers:

```{r formatting-example}
big_number <- 1234567.89

# Without formatting:
# The budget was `r big_number` dollars.

# Better — use format() or formatC():
# The budget was `r formatC(big_number, format = "f", big.mark = ",", digits = 0)` dollars.
```

The budget was `r formatC(big_number, format = "f", big.mark = ",", digits = 0)` dollars.


# Exercise 2.2: Write a Dynamic Paragraph {.unnumbered}

**Hands-on practice (5 minutes):**

Using the data below, write a short paragraph about the results using inline
code. Your paragraph should mention the sample size, mean concentration, and
the percentage above the reference dose.

```{r exercise-inline}
# Simulated BPA urinary concentrations (µg/L)
set.seed(123)
bpa_data <- data.frame(
  participant_id = 1:80,
  bpa_concentration = round(rlnorm(80, meanlog = 0.5, sdlog = 0.9), 2),
  age_group = sample(c("18-39", "40-59", "60+"), 80, replace = TRUE)
)

# Reference dose threshold
reference_dose <- 5.0  # µg/L

# Calculate some statistics you can use inline
n_total <- nrow(bpa_data)
mean_bpa <- round(mean(bpa_data$bpa_concentration), 2)
pct_above <- round(mean(bpa_data$bpa_concentration > reference_dose) * 100, 1)
```

**Write your paragraph below this line** (in the `.Rmd` source), using inline R
code like `` `r knitr::inline_expr('n_total')` `` to insert the values:

<!-- YOUR PARAGRAPH HERE -->
<!-- Example: "We measured BPA in `r knitr::inline_expr('n_total')` participants..." -->



# Recap

In this module you learned:

- **Code chunks** are fenced blocks that run R code during knitting
- **Chunk options** (`echo`, `eval`, `include`, `message`, `warning`) control
  what appears in the output
- **Global options** set defaults for the whole document via
  `knitr::opts_chunk$set()`
- **Inline code** embeds R results directly in text — essential for
  reproducible reporting

---

**Next up:** `03_tables_and_data.Rmd` — making your data look professional.
