---
title: "Module 3: Tables and Data Display"
subtitle: "From Raw Data to Publication-Quality Tables"
author: "NIEHS Training"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
```

# The Problem with Default R Output

When you print a data frame in R Markdown, the default output is plain text — 
functional, but not appropriate for reports or manuscripts.

```{r default-output}
# Simulated environmental monitoring data
air_quality <- data.frame(
  Site = c("Research Triangle", "Durham Central", "Chapel Hill", 
           "Raleigh Downtown", "Morrisville"),
  PM2.5_mean = c(8.2, 9.7, 7.4, 10.1, 8.8),
  PM2.5_max = c(22.1, 31.4, 18.9, 35.7, 24.3),
  Ozone_mean = c(0.042, 0.048, 0.039, 0.051, 0.044),
  Days_AQI_Good = c(312, 287, 325, 271, 298)
)

# Default output — ugly in a report
air_quality
```

Let's fix that.


# Method 1: `knitr::kable()` — Simple and Reliable

`kable()` is the workhorse table function in R Markdown. It is built into the
`knitr` package and works with all output formats (HTML, Word, PDF).

```{r kable-basic}
kable(air_quality, 
      caption = "Table 1. Annual Air Quality Summary for Triangle Region Sites")
```

## Improving `kable()` Output

```{r kable-improved}
kable(air_quality,
      col.names = c("Monitoring Site", "Mean PM₂.₅ (µg/m³)", 
                     "Max PM₂.₅ (µg/m³)", "Mean O₃ (ppm)", 
                     "Days AQI 'Good'"),
      digits = 3,
      align = c("l", "r", "r", "r", "r"),
      caption = "Table 1. Annual Air Quality Summary for Triangle Region Sites")
```


# Method 2: `kableExtra` — Publication-Quality Styling

The `kableExtra` package extends `kable()` with powerful formatting options.
It works best with HTML output.

```{r kableextra-demo}
kable(air_quality,
      col.names = c("Monitoring Site", "Mean PM₂.₅ (µg/m³)", 
                     "Max PM₂.₅ (µg/m³)", "Mean O₃ (ppm)", 
                     "Days AQI 'Good'"),
      digits = 3,
      caption = "Table 1. Annual Air Quality Summary") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(which(air_quality$PM2.5_max > 30), 
           bold = TRUE, color = "#DC2626") %>%
  footnote(general = "Sites exceeding 30 µg/m³ max PM₂.₅ shown in red.")
```

## Key `kable_styling()` Options

| Option | Values | Purpose |
|--------|--------|---------|
| `bootstrap_options` | `"striped"`, `"hover"`, `"condensed"`, `"bordered"` | Visual style |
| `full_width` | `TRUE` / `FALSE` | Stretch to full page width? |
| `position` | `"center"`, `"left"`, `"right"`, `"float_right"` | Table alignment |
| `font_size` | Number (e.g., 12) | Text size in points |

## Adding Row and Column Formatting

```{r kableextra-formatting}
# Study results table
results <- data.frame(
  Exposure = c("PFOA (high)", "PFOA (low)", "PFOS (high)", "PFOS (low)"),
  N = c(245, 251, 238, 258),
  Beta = c(0.42, 0.18, 0.55, 0.22),
  SE = c(0.11, 0.09, 0.13, 0.10),
  P_value = c(0.0002, 0.048, 0.00003, 0.028)
)

results %>%
  mutate(Significance = ifelse(P_value < 0.01, "**", 
                               ifelse(P_value < 0.05, "*", ""))) %>%
  kable(col.names = c("Exposure Group", "N", "β", "SE", "P-value", ""),
        digits = c(0, 0, 2, 2, 4, 0),
        caption = "Table 2. Association Between PFAS Exposure and Outcome") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) %>%
  add_header_above(c(" " = 2, "Regression Estimates" = 3, " " = 1)) %>%
  footnote(symbol = c("p < 0.05", "p < 0.01"))
```


## Grouped Headers

```{r grouped-header}
# Comparison table
comparison <- data.frame(
  Chemical = c("PFOA", "PFOS", "BPA"),
  Serum_N = c(482, 482, 475),
  Serum_Mean = c(3.21, 5.87, NA),
  Urine_N = c(NA, NA, 475),
  Urine_Mean = c(NA, NA, 2.14)
)

kable(comparison,
      col.names = c("Chemical", "N", "Mean (ng/mL)", "N", "Mean (µg/L)"),
      digits = 2,
      caption = "Table 3. Biomarker Concentrations by Matrix") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Serum" = 2, "Urine" = 2))
```


# Method 3: `DT::datatable()` — Interactive Tables

For HTML output, `DT` provides searchable, sortable, and pageable tables.
These are great for exploratory data and appendices.

```{r dt-demo}
library(DT)

# Generate a larger dataset
set.seed(42)
participants <- data.frame(
  ID = paste0("NIEHS-", sprintf("%04d", 1:50)),
  Age = round(rnorm(50, 45, 12)),
  Sex = sample(c("Female", "Male"), 50, replace = TRUE),
  BMI = round(rnorm(50, 27, 4.5), 1),
  PFOA_ng_mL = round(rlnorm(50, 1.2, 0.8), 2),
  PFOS_ng_mL = round(rlnorm(50, 1.5, 0.7), 2),
  Site = sample(c("RTP", "Durham", "Chapel Hill"), 50, replace = TRUE)
)

datatable(participants,
          caption = "Participant Data (searchable and sortable)",
          filter = "top",
          options = list(pageLength = 10,
                         autoWidth = TRUE)) %>%
  formatRound(columns = c("PFOA_ng_mL", "PFOS_ng_mL"), digits = 2)
```

> **Note:** `DT` tables are interactive in HTML output only. They will not
> appear in Word or PDF documents. For those formats, use `kable()` instead.


# Displaying Model Output Cleanly

Raw model output is messy. The `broom` package tidies it for table display.

```{r model-output}
library(broom)

# Fit a simple model
model <- lm(PFOA_ng_mL ~ Age + BMI + Sex, data = participants)

# Raw output (ugly):
# summary(model)

# Tidy output with broom:
tidy(model, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(col.names = c("Term", "Estimate", "SE", "Statistic", 
                       "P-value", "Lower 95% CI", "Upper 95% CI"),
        caption = "Table 4. Linear Regression: PFOA ~ Age + BMI + Sex") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

```{r model-fit}
# Model fit statistics
glance(model) %>%
  select(r.squared, adj.r.squared, sigma, statistic, p.value, nobs) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(col.names = c("R²", "Adj. R²", "σ", "F", "P-value", "N"),
        caption = "Model Fit Statistics") %>%
  kable_styling(bootstrap_options = c("condensed"), full_width = FALSE)
```


# Exercise 3.1: Build a Summary Table {.unnumbered}

**Hands-on practice (5 minutes):**

Using the `participants` data created above, produce a summary table that shows
the mean and SD of `Age`, `BMI`, `PFOA_ng_mL`, and `PFOS_ng_mL` grouped by
`Sex`. Use `kable()` and `kable_styling()` to make it look professional.

Starter code:

```{r exercise-table, eval=FALSE}
# Step 1: Summarize the data
summary_table <- participants %>%
  group_by(Sex) %>%
  summarise(
    N = n(),
    `Mean Age` = round(mean(Age), 1),
    `SD Age` = round(sd(Age), 1),
    `Mean BMI` = round(mean(BMI), 1),
    # Add PFOA and PFOS summaries here...
  )

# Step 2: Display with kable
summary_table %>%
  kable(caption = "YOUR CAPTION HERE") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```


# When to Use Which Table Method

| Method | Best for | Output formats |
|--------|----------|----------------|
| `kable()` | Simple, reliable tables | HTML, Word, PDF |
| `kableExtra` | Styled, publication-quality tables | HTML (best), PDF |
| `DT::datatable()` | Interactive exploration, appendices | HTML only |
| `gt` package | Highly customized tables | HTML, PDF |

---

**Next up:** `04_visualizations.Rmd` — figures and plots in R Markdown.
