---
title: "Module 4: Visualizations and Figure Control"
subtitle: "Publication-Quality Figures in R Markdown"
author: "NIEHS Training"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  dpi = 150
)
library(tidyverse)
```

# Figures in R Markdown

One of the biggest advantages of R Markdown is that figures are generated
directly from your code. No more saving PNGs and pasting them into Word —
the plot IS the code.

# Creating Figures with ggplot2

Let's start with environmental health data and build up a publication-ready
figure.

```{r sample-data}
# Simulated biomonitoring data
set.seed(42)
n <- 200
biomarker_data <- data.frame(
  participant_id = 1:n,
  age = round(rnorm(n, 48, 14)),
  sex = sample(c("Female", "Male"), n, replace = TRUE, prob = c(0.55, 0.45)),
  pfoa_serum = round(rlnorm(n, meanlog = 0.8, sdlog = 0.9), 2),
  cholesterol = round(rnorm(n, 210, 38)),
  site = sample(c("Research Triangle Park", "Durham", "Chapel Hill"), 
                n, replace = TRUE)
)
```

## A Basic Plot

```{r basic-plot}
ggplot(biomarker_data, aes(x = pfoa_serum, y = cholesterol)) +
  geom_point() +
  labs(title = "PFOA Serum Concentration vs. Total Cholesterol",
       x = "PFOA (ng/mL)",
       y = "Total Cholesterol (mg/dL)")
```

## A Publication-Quality Plot

```{r pub-quality}
ggplot(biomarker_data, aes(x = pfoa_serum, y = cholesterol, color = sex)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  scale_x_log10() +
  scale_color_manual(values = c("Female" = "#0D9488", "Male" = "#6366F1")) +
  labs(
    title = "Association Between Serum PFOA and Total Cholesterol",
    subtitle = "Stratified by sex, N = 200",
    x = "Serum PFOA (ng/mL, log scale)",
    y = "Total Cholesterol (mg/dL)",
    color = "Sex",
    caption = "Data: Simulated biomonitoring study | NIEHS Training"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```


# Controlling Figure Size and Resolution

Figure dimensions are set via chunk options. This is critical for getting
figures that look right in your reports and manuscripts.

## Key Figure Chunk Options

| Option | Purpose | Default | Example |
|--------|---------|---------|---------|
| `fig.width` | Width in inches | 7 | `fig.width=8` |
| `fig.height` | Height in inches | 5 | `fig.height=4` |
| `fig.align` | Alignment | `"default"` | `"center"`, `"left"`, `"right"` |
| `dpi` | Resolution (dots per inch) | 72 | `150` for reports, `300` for publication |
| `fig.cap` | Figure caption | — | `"Figure 1. My caption"` |
| `out.width` | Rendered width (CSS/LaTeX) | — | `"80%"`, `"500px"` |

## Same Plot, Different Sizes

Here is a narrow, tall figure:

```{r fig-narrow, fig.width=4, fig.height=6}
ggplot(biomarker_data, aes(x = sex, y = pfoa_serum, fill = sex)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#0D9488", "#6366F1")) +
  scale_y_log10() +
  labs(y = "Serum PFOA (ng/mL)", x = NULL) +
  theme_minimal() +
  theme(legend.position = "none")
```

And here is a wide, short figure:

```{r fig-wide, fig.width=10, fig.height=3}
ggplot(biomarker_data, aes(x = age, y = cholesterol)) +
  geom_point(alpha = 0.4, color = "#475569") +
  geom_smooth(method = "loess", color = "#0D9488") +
  facet_wrap(~site, nrow = 1) +
  labs(x = "Age (years)", y = "Total Cholesterol (mg/dL)") +
  theme_minimal()
```

## Using `out.width` for HTML Scaling

`fig.width` and `fig.height` control the *creation* size. `out.width` controls
how large it *appears* in the rendered document:

```{r scaled-down, fig.width=8, fig.height=6, out.width="60%"}
ggplot(biomarker_data, aes(x = pfoa_serum)) +
  geom_histogram(fill = "#0D9488", color = "white", bins = 30) +
  scale_x_log10() +
  labs(title = "Distribution of Serum PFOA",
       x = "PFOA (ng/mL, log scale)", y = "Count") +
  theme_minimal()
```

This figure was created at 8×6 inches but displayed at 60% width.


# Figure Captions

Use the `fig.cap` chunk option for proper figure captions:

```{r captioned-figure, fig.cap="**Figure 1.** Serum PFOA concentration by monitoring site. Boxes show median and IQR; whiskers extend to 1.5× IQR.", fig.width=7, fig.height=4}
ggplot(biomarker_data, aes(x = site, y = pfoa_serum, fill = site)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_log10() +
  labs(x = NULL, y = "Serum PFOA (ng/mL, log scale)") +
  theme_minimal() +
  theme(legend.position = "none")
```


# Multiple Plots in One Figure

## Using `patchwork`

The `patchwork` package combines separate ggplots into panels:

```{r patchwork-demo, fig.width=10, fig.height=4}
# install.packages("patchwork")  # Uncomment if needed
library(patchwork)

p1 <- ggplot(biomarker_data, aes(x = pfoa_serum)) +
  geom_histogram(fill = "#0D9488", color = "white", bins = 25) +
  scale_x_log10() +
  labs(title = "A", x = "PFOA (ng/mL)", y = "Count") +
  theme_minimal()

p2 <- ggplot(biomarker_data, aes(x = pfoa_serum, y = cholesterol)) +
  geom_point(alpha = 0.4, color = "#475569") +
  geom_smooth(method = "lm", color = "#DC2626") +
  scale_x_log10() +
  labs(title = "B", x = "PFOA (ng/mL)", y = "Cholesterol (mg/dL)") +
  theme_minimal()

p1 + p2
```

## Using `facet_wrap` and `facet_grid`

```{r faceted-plot, fig.width=8, fig.height=5}
ggplot(biomarker_data, aes(x = age, y = cholesterol)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", color = "#0D9488") +
  facet_grid(sex ~ site) +
  labs(
    title = "Age vs. Cholesterol by Site and Sex",
    x = "Age (years)", y = "Total Cholesterol (mg/dL)"
  ) +
  theme_bw(base_size = 10)
```


# Interactive Plots with plotly

For HTML reports, `plotly` turns static ggplots into interactive figures where
readers can hover, zoom, and pan.

```{r plotly-demo, fig.width=7, fig.height=5}
library(plotly)

p <- ggplot(biomarker_data, aes(x = pfoa_serum, y = cholesterol, 
                                 color = sex, text = paste("Age:", age))) +
  geom_point(alpha = 0.6) +
  scale_x_log10() +
  scale_color_manual(values = c("#0D9488", "#6366F1")) +
  labs(x = "PFOA (ng/mL)", y = "Cholesterol (mg/dL)") +
  theme_minimal()

ggplotly(p, tooltip = c("text", "x", "y"))
```

> **Hover over the points** to see individual data. Use the toolbar in the
> upper right to zoom, pan, and download the image.

> **Note:** `plotly` figures only work in HTML output. They will not appear in
> Word or PDF.


# Saving Figures

Sometimes you need a figure file for a manuscript submission. Use `ggsave()`:

```{r save-figure, eval=FALSE}
# Save the last plot
ggsave("output/figures/pfoa_cholesterol.png", 
       width = 7, height = 5, dpi = 300)

# Save a specific plot object
ggsave("output/figures/panel_figure.png", 
       plot = p1 + p2, 
       width = 10, height = 4, dpi = 300)

# For journals that require TIFF:
ggsave("output/figures/figure1.tiff", 
       width = 7, height = 5, dpi = 300, compression = "lzw")
```


# Exercise 4.1: Create a Multi-Panel Figure {.unnumbered}

**Hands-on practice (5 minutes):**

Using `biomarker_data`, create a figure with two panels side by side:

- **Panel A:** A histogram of `cholesterol` values
- **Panel B:** A scatterplot of `age` vs. `cholesterol` colored by `sex`

Use `patchwork` to combine them. Set `fig.width=10, fig.height=4` in the
chunk options. Add a figure caption with `fig.cap`.

```{r exercise-figure, eval=FALSE, fig.width=10, fig.height=4}
# Your code here
panel_a <- ggplot(biomarker_data, aes(x = cholesterol)) +
  # ... add layers ...
  NULL

panel_b <- ggplot(biomarker_data, aes(x = age, y = cholesterol)) +
  # ... add layers ...
  NULL

# Combine with patchwork
# panel_a + panel_b
```


---

**Next up:** `05_scientific_writing.Rmd` — equations, citations, and parameters.
