---
title: "Module 6: Output Formats, Themes, and Templates"
subtitle: "HTML, Word, PDF, and Beyond"
author: "NIEHS Training"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
```

# Output Formats Overview

The `output` field in your YAML header determines the format of the knitted
document. R Markdown supports many output types.

## The Big Three

| Format | YAML value | Best for |
|--------|-----------|----------|
| HTML | `html_document` | Interactive reports, sharing online |
| Word | `word_document` | Manuscripts, institutional reports |
| PDF | `pdf_document` | Formal publications (requires LaTeX) |


# HTML Output Options

HTML is the most flexible output format. This entire training uses it.

## Themes

R Markdown includes several built-in Bootstrap themes. Change the `theme`
field in the YAML header:

Available themes: `default`, `bootstrap`, `cerulean`, `cosmo`, `darkly`,
`flatly`, `journal`, `lumen`, `paper`, `readable`, `sandstone`, `simplex`,
`spacelab`, `united`, `yeti`

````yaml
output:
  html_document:
    theme: flatly    # Try changing this to "journal" or "cosmo"
````

## Syntax Highlighting

Control how code chunks are colored with the `highlight` option:

Available styles: `default`, `tango`, `pygments`, `kate`, `monochrome`,
`espresso`, `zenburn`, `haddock`, `breezedark`

````yaml
output:
  html_document:
    highlight: tango
````

## Table of Contents

```yaml
output:
  html_document:
    toc: true            # Enable table of contents
    toc_float: true      # Float it on the left side
    toc_depth: 3         # How many heading levels to include
    number_sections: true # Number the sections
```

The floating TOC on the left side of this document is configured exactly this
way.

## Code Folding

The `code_folding` option lets readers show/hide code:

````yaml
output:
  html_document:
    code_folding: hide    # Code hidden by default, click to show
    # code_folding: show  # Code shown by default, click to hide
````

**Notice** the "Code" buttons on the chunks in this document — they let you
toggle code visibility. This file uses `code_folding: hide`.

## Data Frame Printing

Control how data frames appear by default:

````yaml
output:
  html_document:
    df_print: kable     # Use kable formatting automatically
    # df_print: paged   # Paginated HTML tables
    # df_print: default # Plain text
````

```{r df-print-demo}
# With df_print: kable, this automatically renders as a styled table
head(mtcars[, 1:6])
```


# Word Output

Word output is essential for many NIEHS workflows — manuscript drafts,
institutional reports, and documents for non-R-using collaborators.

## Basic Word Output

````yaml
---
title: "My Report"
output: word_document
---
````

## Word with a Reference Document (Template)

You can apply your institution's formatting by providing a Word template:

````yaml
output:
  word_document:
    reference_docx: "templates/niehs_template.docx"
````

**How to create a template:**

1. Knit a basic `.Rmd` to Word
2. Open the `.docx` file and modify the styles (heading fonts, colors, margins)
3. Save this styled file as your template
4. Point to it with `reference_docx`

## Word Output Limitations

- No interactive features (plotly, DT, tabsets)
- No floating TOC
- `kableExtra` styling is limited — stick with basic `kable()` for Word
- Code folding is not available

## Conditional Output

You can write code that behaves differently depending on the output format:

```{r conditional-output, eval=FALSE}
if (knitr::is_html_output()) {
  DT::datatable(my_data)  # Interactive table for HTML
} else {
  kable(my_data)           # Static table for Word/PDF
}
```


# PDF Output

PDF output uses LaTeX under the hood and produces beautifully typeset documents.

## Basic PDF Output

````yaml
---
title: "My Report"
output: pdf_document
---
````

## Posit Cloud and PDF

To knit to PDF on Posit Cloud, you need TinyTeX installed:

```{r install-tinytex-reminder, eval=FALSE}
install.packages("tinytex")
tinytex::install_tinytex()
```

> **For this workshop, we recommend HTML output.** PDF is optional and can be
> set up later. If you encounter LaTeX errors, switch back to `html_document`.

## PDF Options

````yaml
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    latex_engine: xelatex   # Better Unicode support
    keep_tex: false          # Set TRUE to debug LaTeX issues
````


# Multiple Output Formats

You can specify multiple output formats in one YAML header:

````yaml
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
  word_document:
    reference_docx: "template.docx"
  pdf_document:
    toc: true
````

The **Knit** button dropdown lets you choose which format to render.


# Useful YAML Options Reference

Here is a comprehensive YAML configuration for a polished HTML report:

````yaml
---
title: "Dose-Response Analysis of PFAS Exposure in the Triangle Region"
subtitle: "Supplementary Analysis Report"
author:
  - "Jane Smith, Ph.D. — Biostatistics Branch, NIEHS"
  - "John Doe, Ph.D. — Epidemiology Branch, NIEHS"
date: "Report generated: `r knitr::inline_expr('format(Sys.Date(), \"%B %d, %Y\")')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    df_print: kable
    fig_caption: true
params:
  data_file: "data/pfas_2024.csv"
  alpha: 0.05
---
````


# Custom CSS (Advanced)

For full control over appearance, you can add custom CSS:

````yaml
output:
  html_document:
    css: custom_styles.css
````

Or embed CSS directly in your `.Rmd`:

```{css custom-css, echo=TRUE}
/* Make the title green */
.title {
  color: #1B5E4B;
}

/* Style block quotes */
blockquote {
  border-left: 4px solid #0D9488;
  padding-left: 12px;
  color: #475569;
}
```


# Exercise 6.1: Experiment with Output Options {.unnumbered}

**Hands-on practice (3 minutes):**

1. In this file's YAML, change `theme: flatly` to `theme: journal`
2. Change `code_folding: hide` to `code_folding: show`
3. Knit and observe the differences
4. Try `theme: darkly` for a dark mode experience
5. Change back when you're done

---

**Next up:** `07_capstone_exercise.Rmd` — put everything together!
