---
title: "Module 7 — Capstone Exercise"
subtitle: "Build a Complete Environmental Health Analysis Report"
author: "YOUR NAME HERE"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    df_print: kable
params:
  seed: 42
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  dpi = 150
)
```

<!--
=====================================================================
CAPSTONE EXERCISE — 20 minutes

In this exercise you will build a complete, reproducible analysis report
by filling in the sections below. Each section has instructions in HTML
comments (like this one) telling you what to add.

Skills you will practice:
  - YAML configuration
  - Loading packages and data
  - Inline code for dynamic reporting
  - kable/kableExtra tables
  - ggplot2 figures with chunk options
  - LaTeX equations
  - Tabbed sections

START by changing the author field in the YAML header to your name!
=====================================================================
-->


# Introduction

<!-- 
INSTRUCTIONS: Write 2-3 sentences describing the purpose of this analysis.
Mention that you are examining the relationship between an environmental
exposure (PM2.5 air pollution) and a health outcome (respiratory function
measured as FEV1) in a cohort near Research Triangle Park.

Use **bold** for the chemical name and *italics* for the study design.
-->

YOUR INTRODUCTION HERE


# Methods

## Study Population

```{r load-data}
# ---------------------------------------------------------------
# DO NOT MODIFY THIS CHUNK — it generates the study data
# ---------------------------------------------------------------
set.seed(params$seed)
n <- 250

study_data <- data.frame(
  participant_id = paste0("RTP-", sprintf("%04d", 1:n)),
  age = round(rnorm(n, mean = 52, sd = 14)),
  sex = sample(c("Female", "Male"), n, replace = TRUE, prob = c(0.54, 0.46)),
  bmi = round(rnorm(n, mean = 28.1, sd = 5.2), 1),
  smoking_status = sample(c("Never", "Former", "Current"), n, 
                          replace = TRUE, prob = c(0.50, 0.30, 0.20)),
  pm25_annual = round(rnorm(n, mean = 9.5, sd = 2.8), 1),
  fev1_liters = NA
)

# Generate FEV1 with a true relationship to PM2.5, age, smoking
study_data$fev1_liters <- round(
  3.8 
  - 0.025 * study_data$age 
  - 0.04 * study_data$pm25_annual 
  + ifelse(study_data$sex == "Male", 0.6, 0) 
  - ifelse(study_data$smoking_status == "Current", 0.35, 
           ifelse(study_data$smoking_status == "Former", 0.15, 0))
  + rnorm(n, 0, 0.4),
  2
)

# Ensure reasonable values
study_data$fev1_liters <- pmax(study_data$fev1_liters, 1.0)
study_data$age <- pmax(study_data$age, 25)
```

<!--
INSTRUCTIONS: Write a sentence using INLINE CODE to report:
  - The total number of participants (use `nrow(study_data)`)
  - The percentage female
  - The age range

Example format:
"The study enrolled `r knitr::inline_expr('nrow(study_data)')` participants..."
-->

YOUR STUDY POPULATION DESCRIPTION HERE

## Statistical Analysis

<!--
INSTRUCTIONS: Describe the statistical approach. Include:
  1. A sentence about using linear regression
  2. The regression equation using LaTeX display math ($$...$$):
     FEV1 = β₀ + β₁(PM2.5) + β₂(Age) + β₃(Sex) + β₄(Smoking) + ε
  3. Mention that α = 0.05
-->

YOUR METHODS DESCRIPTION HERE


# Results

## Descriptive Statistics

<!--
INSTRUCTIONS: Create a summary table using kable() and kableExtra.
  - Group by sex
  - Show N, mean age, mean BMI, mean PM2.5, and mean FEV1
  - Add a caption
  - Use kable_styling with "striped" and "hover"
  - Add a grouped header row
-->

```{r descriptive-table}
library(knitr)
library(kableExtra)
library(tidyverse)

# YOUR TABLE CODE HERE
# Hint: Use group_by(sex) %>% summarise(...)
```

## Exposure Distribution {.tabset}

<!--
INSTRUCTIONS: Create a tabbed section with two tabs:
  - Tab 1: "Histogram" — histogram of PM2.5 values
  - Tab 2: "By Smoking Status" — boxplot of PM2.5 by smoking_status
  
Remember: tabs are created by putting content under ### headers
when the parent ## header has {.tabset}
-->

### Histogram

```{r pm25-histogram, fig.cap="**Figure 1.** Distribution of annual PM₂.₅ exposure (µg/m³)."}
# YOUR HISTOGRAM CODE HERE
# Hint: ggplot(study_data, aes(x = pm25_annual)) + geom_histogram(...)
```

### By Smoking Status

```{r pm25-boxplot, fig.cap="**Figure 2.** PM₂.₅ exposure by smoking status."}
# YOUR BOXPLOT CODE HERE
```

##

## Primary Analysis

<!--
INSTRUCTIONS: 
  1. Fit the linear model: fev1_liters ~ pm25_annual + age + sex + smoking_status
  2. Use broom::tidy() to extract results
  3. Display results in a kable table with proper column names
  4. Write a sentence below the table using INLINE CODE to report:
     - The PM2.5 coefficient (β₁)
     - Its p-value
     - Whether it is statistically significant
-->

```{r regression-model}
library(broom)

# YOUR MODEL CODE HERE
# model <- lm(fev1_liters ~ pm25_annual + age + sex + smoking_status, 
#              data = study_data)
# results <- tidy(model, conf.int = TRUE)
```

## Visualization

<!--
INSTRUCTIONS: Create a scatterplot of pm25_annual vs fev1_liters
  - Color by sex
  - Add a regression line (geom_smooth with method = "lm")
  - Use appropriate labels and a theme
  - Set fig.width=8, fig.height=5
  - Add a fig.cap
-->

```{r main-figure, fig.width=8, fig.height=5}
# YOUR SCATTERPLOT CODE HERE
```


# Discussion

<!--
INSTRUCTIONS: Write 2-3 sentences interpreting the results.
  - What direction was the PM2.5-FEV1 association?
  - Was it statistically significant?
  - What confounders were adjusted for?
  - Mention one limitation (e.g., cross-sectional design)

Use inline code to reference at least one number from the analysis.
-->

YOUR DISCUSSION HERE


# Session Info

This section documents the R environment used to produce this report, supporting
reproducibility.

```{r session-info}
sessionInfo()
```

---

<div style="padding: 15px; background-color: #F0FDF4; border-left: 4px solid #0D9488; margin-top: 20px;">

**Congratulations!** If you completed this capstone, you have practiced:

- ✅ Configuring the YAML header (title, author, output options, parameters)
- ✅ Writing narrative text with Markdown formatting
- ✅ Using code chunks with appropriate chunk options
- ✅ Reporting statistics with inline R code
- ✅ Creating publication-quality tables with `kable()` and `kableExtra`
- ✅ Making figures with `ggplot2` and controlling their appearance
- ✅ Writing LaTeX equations
- ✅ Using tabbed sections
- ✅ Hiding code with `code_folding`

**You are now equipped to write reproducible analysis reports with R Markdown!**

</div>
