---
title: "R Markdown Training — Exercise Solutions"
subtitle: "NIEHS Workshop — Instructor Reference"
author: "NIEHS Training"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
library(broom)
```

> **Instructor-only document.** Do not distribute to participants before the
> workshop. These are the completed solutions for all exercises.

---

# Exercise 1.1: Edit the YAML (Module 1)

**Task:** Change the `author` field to your name, then knit.

**Solution:** Simply change the YAML header line:

```yaml
---
title: "Module 1: R Markdown Basics"
author: "Jane Smith, NIEHS"    # ← changed from "NIEHS Training"
date: "`r knitr::inline_expr('Sys.Date()')`"
output: html_document
---
```

No code to run — this exercises YAML editing and re-knitting.

---

# Exercise 1.2: Create Your First R Markdown Document (Module 1)

**Task:** Create a new `.Rmd` via File → New → R Markdown, modify it with bold,
italic, and a numbered list, then knit.

**Solution:** After creating the default template, replace the body with
something like:

```markdown
# About This Document

This is my **first R Markdown document**. I created it during the
*NIEHS R Markdown Workshop*.

## What I Want to Learn Today

1. How to create reproducible reports
2. How to embed R output in documents
3. How to make publication-quality tables and figures
```

Then knit to HTML.

---

# Exercise 2.1: Practice Chunk Options (Module 2)

**Task:** Change `echo=TRUE` to `echo=FALSE` on the PFAS half-life chunk.

**Solution:**

```{r exercise-2-1, echo=FALSE}
chemicals <- c("PFOA", "PFOS", "PFNA", "PFHxS", "GenX")
half_life_years <- c(3.8, 5.4, 2.5, 8.5, NA)

data.frame(
  Chemical = chemicals,
  `Half-life (years)` = half_life_years,
  check.names = FALSE
)
```

The table appears but the code is hidden. Participants should see only the
output when they knit.

---

# Exercise 2.2: Write a Dynamic Paragraph (Module 2)

**Task:** Using the BPA data, write a paragraph with inline R code.

**Solution:**

```{r exercise-2-2-data}
set.seed(123)
bpa_data <- data.frame(
  participant_id = 1:80,
  bpa_concentration = round(rlnorm(80, meanlog = 0.5, sdlog = 0.9), 2),
  age_group = sample(c("18-39", "40-59", "60+"), 80, replace = TRUE)
)

reference_dose <- 5.0

n_total <- nrow(bpa_data)
mean_bpa <- round(mean(bpa_data$bpa_concentration), 2)
median_bpa <- round(median(bpa_data$bpa_concentration), 2)
pct_above <- round(mean(bpa_data$bpa_concentration > reference_dose) * 100, 1)
max_bpa <- round(max(bpa_data$bpa_concentration), 2)
```

**The paragraph (written in the `.Rmd` source with inline code):**

We measured urinary BPA concentrations in `r n_total` participants. The mean
concentration was `r mean_bpa` µg/L (median: `r median_bpa` µg/L), with a
maximum of `r max_bpa` µg/L. Of the participants, `r pct_above`% had
concentrations exceeding the `r reference_dose` µg/L reference dose.

**The raw Markdown for the above (what participants should type):**

````
We measured urinary BPA concentrations in `r knitr::inline_expr('n_total')` participants. The mean
concentration was `r knitr::inline_expr('mean_bpa')` µg/L (median: `r knitr::inline_expr('median_bpa')` µg/L), with a
maximum of `r knitr::inline_expr('max_bpa')` µg/L. Of the participants, `r knitr::inline_expr('pct_above')`% had
concentrations exceeding the `r knitr::inline_expr('reference_dose')` µg/L reference dose.
````

---

# Exercise 3.1: Build a Summary Table (Module 3)

**Task:** Create a summary table of Age, BMI, PFOA, and PFOS grouped by Sex.

**Solution:**

```{r exercise-3-1}
# Recreate the participants data from Module 3
set.seed(42)
participants <- data.frame(
  ID = paste0("NIEHS-", sprintf("%04d", 1:50)),
  Age = round(rnorm(50, 45, 12)),
  Sex = sample(c("Female", "Male"), 50, replace = TRUE),
  BMI = round(rnorm(50, 27, 4.5), 1),
  PFOA_ng_mL = round(rlnorm(50, 1.2, 0.8), 2),
  PFOS_ng_mL = round(rlnorm(50, 1.5, 0.7), 2),
  Site = sample(c("RTP", "Durham", "Chapel Hill"), 50, replace = TRUE)
)

# Step 1: Summarize the data
summary_table <- participants %>%
  group_by(Sex) %>%
  summarise(
    N = n(),
    `Mean Age` = round(mean(Age), 1),
    `SD Age` = round(sd(Age), 1),
    `Mean BMI` = round(mean(BMI), 1),
    `SD BMI` = round(sd(BMI), 1),
    `Mean PFOA (ng/mL)` = round(mean(PFOA_ng_mL), 2),
    `SD PFOA` = round(sd(PFOA_ng_mL), 2),
    `Mean PFOS (ng/mL)` = round(mean(PFOS_ng_mL), 2),
    `SD PFOS` = round(sd(PFOS_ng_mL), 2)
  )

# Step 2: Display with kable
summary_table %>%
  kable(caption = "Table 1. Participant Characteristics by Sex") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) %>%
  add_header_above(c(" " = 2, "Age" = 2, "BMI" = 2, "PFOA" = 2, "PFOS" = 2))
```

---

# Exercise 4.1: Create a Multi-Panel Figure (Module 4)

**Task:** Two panels — histogram of cholesterol + scatterplot of age vs
cholesterol colored by sex — combined with `patchwork`.

**Solution:**

```{r exercise-4-1, fig.width=10, fig.height=4, fig.cap="**Figure 1.** (A) Distribution of total cholesterol. (B) Age vs. cholesterol stratified by sex with linear regression lines."}
library(patchwork)

# Recreate biomarker data from Module 4
set.seed(42)
n <- 200
biomarker_data <- data.frame(
  age = round(rnorm(n, 48, 14)),
  sex = sample(c("Female", "Male"), n, replace = TRUE, prob = c(0.55, 0.45)),
  pfoa_serum = round(rlnorm(n, 0.8, 0.9), 2),
  cholesterol = round(rnorm(n, 210, 38)),
  site = sample(c("RTP", "Durham", "Chapel Hill"), n, replace = TRUE)
)

panel_a <- ggplot(biomarker_data, aes(x = cholesterol)) +
  geom_histogram(fill = "#0D9488", color = "white", bins = 25) +
  labs(title = "A",
       x = "Total Cholesterol (mg/dL)",
       y = "Count") +
  theme_minimal()

panel_b <- ggplot(biomarker_data, aes(x = age, y = cholesterol, color = sex)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c("Female" = "#0D9488", "Male" = "#6366F1")) +
  labs(title = "B",
       x = "Age (years)",
       y = "Total Cholesterol (mg/dL)",
       color = "Sex") +
  theme_minimal() +
  theme(legend.position = "bottom")

panel_a + panel_b
```

---

# Exercise 5.1: Write an Equation and Use Parameters (Module 5)

**Task:** (1) Write the odds ratio equation. (2) Change parameters and re-knit.

**Solution:**

### Part 1: The equation

```markdown
$$OR = \frac{a / b}{c / d} = \frac{a \cdot d}{b \cdot c}$$
```

Renders as:

$$OR = \frac{a / b}{c / d} = \frac{a \cdot d}{b \cdot c}$$

### Part 2: Change parameters

In the YAML header of `05_scientific_writing.Rmd`, change:

```yaml
params:
  chemical: "PFOS"       # was "PFOA"
  threshold: 6.0         # was 4.0
  study_year: 2025       # was 2024
```

Re-knit — the title, histogram, threshold line, and all inline statistics
update automatically.

---

# Exercise 6.1: Experiment with Output Options (Module 6)

**Task:** Change theme and code_folding settings, knit to see differences.

**Solution:** No code needed — this is a YAML editing exercise:

```yaml
# Try each of these one at a time and re-knit:

# Change 1: Different theme
output:
  html_document:
    theme: journal        # was flatly — serif font, different feel

# Change 2: Show code by default
    code_folding: show    # was hide

# Change 3: Dark mode
    theme: darkly         # dark background, light text

# Remember to change back when done:
    theme: flatly
    code_folding: hide
```

---

# Capstone Exercise Solutions (Module 7)

This is the complete, filled-in version of `07_capstone_exercise.Rmd`.

## YAML Header

```yaml
---
title: "Module 7 — Capstone Exercise"
subtitle: "Build a Complete Environmental Health Analysis Report"
author: "Jane Smith, NIEHS"
date: "`r knitr::inline_expr('format(Sys.Date(), \"%B %d, %Y\")')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
    df_print: kable
params:
  seed: 42
---
```

## Introduction (Solution)

This report examines the association between **PM₂.₅** air pollution exposure
and respiratory function (measured as FEV₁) in a *cross-sectional cohort study*
near Research Triangle Park, North Carolina. Ambient PM₂.₅ is a key pollutant
of concern for the NIEHS, and understanding its health impacts supports
evidence-based air quality standards.

## Study Population (Solution)

```{r capstone-data}
set.seed(42)
n <- 250

study_data <- data.frame(
  participant_id = paste0("RTP-", sprintf("%04d", 1:n)),
  age = round(rnorm(n, mean = 52, sd = 14)),
  sex = sample(c("Female", "Male"), n, replace = TRUE, prob = c(0.54, 0.46)),
  bmi = round(rnorm(n, mean = 28.1, sd = 5.2), 1),
  smoking_status = sample(c("Never", "Former", "Current"), n,
                          replace = TRUE, prob = c(0.50, 0.30, 0.20)),
  pm25_annual = round(rnorm(n, mean = 9.5, sd = 2.8), 1),
  fev1_liters = NA
)

study_data$fev1_liters <- round(
  3.8 - 0.025 * study_data$age - 0.04 * study_data$pm25_annual +
    ifelse(study_data$sex == "Male", 0.6, 0) -
    ifelse(study_data$smoking_status == "Current", 0.35,
           ifelse(study_data$smoking_status == "Former", 0.15, 0)) +
    rnorm(n, 0, 0.4), 2)
study_data$fev1_liters <- pmax(study_data$fev1_liters, 1.0)
study_data$age <- pmax(study_data$age, 25)

# Pre-compute values for inline code
n_total <- nrow(study_data)
pct_female <- round(mean(study_data$sex == "Female") * 100, 1)
age_range <- range(study_data$age)
mean_age <- round(mean(study_data$age), 1)
```

The study enrolled `r n_total` participants (`r pct_female`% female) with ages
ranging from `r age_range[1]` to `r age_range[2]` years (mean:
`r mean_age` years).

## Statistical Analysis (Solution)

We used **multiple linear regression** to model the association between annual
PM₂.₅ exposure and FEV₁, adjusting for age, sex, and smoking status. The
model is specified as:

$$\text{FEV}_1 = \beta_0 + \beta_1(\text{PM}_{2.5}) + \beta_2(\text{Age}) + \beta_3(\text{Sex}) + \beta_4(\text{Smoking}) + \epsilon$$

Statistical significance was assessed at $\alpha = 0.05$.

## Descriptive Statistics (Solution)

```{r capstone-descriptive}
desc_table <- study_data %>%
  group_by(sex) %>%
  summarise(
    N = n(),
    `Mean Age (yr)` = round(mean(age), 1),
    `SD Age` = round(sd(age), 1),
    `Mean BMI (kg/m²)` = round(mean(bmi), 1),
    `SD BMI` = round(sd(bmi), 1),
    `Mean PM₂.₅ (µg/m³)` = round(mean(pm25_annual), 1),
    `SD PM₂.₅` = round(sd(pm25_annual), 1),
    `Mean FEV₁ (L)` = round(mean(fev1_liters), 2),
    `SD FEV₁` = round(sd(fev1_liters), 2)
  )

desc_table %>%
  kable(caption = "Table 1. Participant Characteristics by Sex") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  add_header_above(c(" " = 2, "Age" = 2, "BMI" = 2, "PM₂.₅" = 2, "FEV₁" = 2))
```

## Exposure Distribution (Solution) {.tabset}

### Histogram

```{r capstone-histogram, fig.cap="**Figure 1.** Distribution of annual PM₂.₅ exposure (µg/m³).", fig.width=7, fig.height=4}
ggplot(study_data, aes(x = pm25_annual)) +
  geom_histogram(fill = "#3182CE", color = "white", bins = 25) +
  labs(title = "Distribution of Annual PM₂.₅ Exposure",
       x = "PM₂.₅ (µg/m³)",
       y = "Count") +
  theme_minimal()
```

### By Smoking Status

```{r capstone-boxplot, fig.cap="**Figure 2.** PM₂.₅ exposure by smoking status.", fig.width=7, fig.height=4}
ggplot(study_data, aes(x = smoking_status, y = pm25_annual,
                        fill = smoking_status)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "PM₂.₅ Exposure by Smoking Status",
       x = "Smoking Status",
       y = "PM₂.₅ (µg/m³)") +
  theme_minimal() +
  theme(legend.position = "none")
```

###

## Primary Analysis (Solution)

```{r capstone-model}
model <- lm(fev1_liters ~ pm25_annual + age + sex + smoking_status,
            data = study_data)

model_results <- tidy(model, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

model_results %>%
  kable(
    col.names = c("Term", "Estimate", "SE", "Statistic",
                  "P-value", "Lower 95% CI", "Upper 95% CI"),
    caption = "Table 2. Linear Regression Results: FEV₁ ~ PM₂.₅ + Covariates"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# Extract PM2.5 coefficient for inline code
pm25_coef <- model_results %>% filter(term == "pm25_annual")
pm25_beta <- pm25_coef$estimate
pm25_p <- pm25_coef$p.value
pm25_sig <- ifelse(pm25_p < 0.05, "statistically significant", "not statistically significant")
```

The PM₂.₅ coefficient was $\beta$ = `r pm25_beta` (p = `r pm25_p`), which is
`r pm25_sig` at the $\alpha$ = 0.05 level. This suggests that each 1 µg/m³
increase in annual PM₂.₅ exposure is associated with a
`r abs(pm25_beta)` L decrease in FEV₁, after adjusting for age, sex, and
smoking status.

## Visualization (Solution)

```{r capstone-figure, fig.width=8, fig.height=5, fig.cap="**Figure 3.** Association between annual PM₂.₅ exposure and FEV₁, stratified by sex."}
ggplot(study_data, aes(x = pm25_annual, y = fev1_liters, color = sex)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  scale_color_manual(values = c("Female" = "#3182CE", "Male" = "#805AD5")) +
  labs(
    title = "PM₂.₅ Exposure and Respiratory Function",
    subtitle = paste("N =", n_total, "participants, adjusted R² =",
                     round(summary(model)$adj.r.squared, 3)),
    x = "Annual PM₂.₅ (µg/m³)",
    y = "FEV₁ (liters)",
    color = "Sex",
    caption = "Data: Simulated RTP cohort | NIEHS Training"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Discussion (Solution)

Our analysis of `r n_total` participants found a `r pm25_sig` negative
association between annual PM₂.₅ exposure and FEV₁ ($\beta$ = `r pm25_beta`,
p = `r pm25_p`). After adjusting for age, sex, and smoking status, the model
explained `r round(summary(model)$adj.r.squared * 100, 1)`% of the variance
in FEV₁.

Male participants had higher FEV₁ values, and current smokers had the lowest
values, both consistent with well-established respiratory physiology. The
negative PM₂.₅ association aligns with prior epidemiological evidence linking
particulate matter exposure to decreased lung function.

**Limitations** include the cross-sectional design, which precludes causal
inference, and the use of area-level PM₂.₅ averages rather than personal
exposure monitoring. Future work should incorporate longitudinal follow-up and
individual-level exposure assessment.

## Session Info

```{r}
sessionInfo()
```
