---
title: "Module 3: Tables & Data Display"
subtitle: "From Raw Data to Publication-Quality Tables"
author: "NIEHS Training"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    css: rmd_slides_style.css
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "", fig.height = 2.8, fig.width = 6)
library(tidyverse); library(knitr); library(kableExtra)
```

# The Problem {.segue}

## Default R Output is Ugly

```{r default-output}
air_quality <- data.frame(
  Site = c("RTP", "Durham", "Chapel Hill", "Raleigh"),
  PM2.5_mean = c(8.2, 9.7, 7.4, 10.1),
  PM2.5_max = c(22.1, 31.4, 18.9, 35.7),
  Days_Good = c(312, 287, 325, 271)
)
air_quality  # raw output — fine for console, ugly in a report
```

Fine for exploration. Not acceptable for a report.

# Method 1: `knitr::kable()` {.segue}

## Simple and Reliable

```{r kable-basic}
kable(air_quality,
      col.names = c("Site", "Mean PM₂.₅", "Max PM₂.₅", "Good AQI Days"),
      digits = 1,
      caption = "Table 1. Air Quality Summary")
```

Works with **all** output formats: HTML, Word, PDF.

# Method 2: `kableExtra` {.segue}

## Publication-Quality Styling

```{r kableextra-demo}
kable(air_quality,
      col.names = c("Site", "Mean PM₂.₅", "Max PM₂.₅", "Good Days"),
      digits = 1, caption = "Annual Air Quality Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "center") %>%
  row_spec(which(air_quality$PM2.5_max > 30), 
           bold = TRUE, color = "#DC2626") %>%
  footnote(general = "Sites with max PM₂.₅ > 30 µg/m³ in red.")
```

## Key `kable_styling()` Options

| Option | Values | Purpose |
|--------|--------|---------|
| `bootstrap_options` | `"striped"`, `"hover"`, `"condensed"` | Visual style |
| `full_width` | `TRUE`/`FALSE` | Stretch to page? |
| `position` | `"center"`, `"left"`, `"float_right"` | Alignment |
| `font_size` | Number | Text size |

## Grouped Headers

```{r grouped}
results <- data.frame(
  Term = c("PFOA (high)", "PFOA (low)", "PFOS (high)"),
  N = c(245, 251, 238),
  Beta = c(0.42, 0.18, 0.55), SE = c(0.11, 0.09, 0.13),
  P = c(0.0002, 0.048, 0.00003))
kable(results, digits = c(0,0,2,2,4),
      col.names = c("Exposure", "N", "β", "SE", "P-value")) %>%
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = FALSE) %>%
  add_header_above(c(" " = 2, "Estimates" = 3))
```

# Method 3: `DT::datatable()` {.segue}

## Interactive Tables (HTML Only)

```{r dt-demo}
library(DT)
set.seed(42)
participants <- data.frame(
  ID = paste0("NIEHS-", sprintf("%03d", 1:20)),
  Age = round(rnorm(20, 45, 12)),
  PFOA = round(rlnorm(20, 1.2, 0.8), 2),
  Site = sample(c("RTP", "Durham"), 20, replace = TRUE))
datatable(participants, filter = "top",
          options = list(pageLength = 4, dom = 'ftp',
                         scrollY = "200px")) %>%
  formatRound("PFOA", 2)
```

<div class="notes">
DT tables are searchable, sortable, and pageable. Great for appendices.
Only works in HTML — not Word or PDF.
</div>

# Model Output {.segue}

## `broom` Makes Models Tidy

```{r broom-demo}
library(broom)
model <- lm(PFOA ~ Age + Site, data = participants)

tidy(model, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(col.names = c("Term", "Est.", "SE", "Stat", 
                       "P", "Lower CI", "Upper CI")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

## When to Use Which Method

| Method | Best for | Formats |
|--------|----------|---------|
| `kable()` | Simple reliable tables | HTML, Word, PDF |
| `kableExtra` | Publication-quality styling | HTML (best), PDF |
| `DT::datatable()` | Interactive exploration | HTML only |
| `gt` package | Highly customized | HTML, PDF |

## Exercise 3.1: Summary Table

Using the `participants` data, create a table showing mean and SD of
Age and PFOA by Site. Use `kable()` + `kable_styling()`.

```r
participants %>%
  group_by(Site) %>%
  summarise(N = n(),
            `Mean Age` = round(mean(Age), 1),
            `Mean PFOA` = round(mean(PFOA), 2)) %>%
  kable(caption = "Summary by Site") %>%
  kable_styling(bootstrap_options = c("striped"), 
                full_width = FALSE)
```

**Time: 5 minutes**
