---
title: "Module 2: Code Chunks & Inline Code"
subtitle: "Controlling What Runs and What Readers See"
author: "NIEHS Training"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    css: rmd_slides_style.css
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "", fig.height = 2.8, fig.width = 6)
library(tidyverse)
```

# Code Chunks {.segue}

## What is a Code Chunk?

A fenced block of R code between `` ```{r} `` and `` ``` ``:

````
```{r chunk-name}`r ''`
# Your R code here
mean(c(1, 2, 3, 4, 5))
```
````

**Three ways to insert a chunk:**

1. Keyboard: `Ctrl+Alt+I` (Cmd+Option+I on Mac)
2. Menu: Code → Insert Chunk
3. Type the fences manually

## Naming Chunks

```
```{r load-exposure-data}`r ''`
exposure <- read.csv("data/exposure.csv")
```
```

Good names help you:

- **Navigate** via the document outline
- **Find errors** ("Error in chunk 'load-exposure-data'")
- **Cache** results efficiently

> Chunk names must be **unique** within a document.

## Running Chunks

| Action | How |
|--------|-----|
| Run current chunk | Green ▶ button, or `Ctrl+Shift+Enter` |
| Run all chunks above | ↓ arrow on chunk toolbar |
| Run a single line | Cursor on line → `Ctrl+Enter` |
| Knit entire document | Knit button or `Ctrl+Shift+K` |

# Chunk Options {.segue}

## The Most Important Options

Options go inside `{r}`, separated by commas:

````
```{r my-chunk, echo=FALSE, fig.width=8}`r ''`
```
````

## `echo` — Show or Hide Code

**`echo=TRUE`** (default) — readers see code AND output:

```{r echo-true-demo}
result <- 2 + 2
result
```

**`echo=FALSE`** — readers see ONLY the output:

```{r echo-false-demo, echo=FALSE}
cat("This result appeared without showing the code.")
```

> Use `echo=FALSE` for final reports where readers want results, not code.

## `eval` — Run or Skip Code

**`eval=FALSE`** — code is displayed but NOT executed:

````
```{r, eval=FALSE}`r ''`
install.packages("some_package")  # Won't actually run
```
````

Useful for showing example code that shouldn't execute during knitting.

## `include` — Show Anything at All?

**`include=FALSE`** — code runs silently, nothing shown:

```{r include-demo, include=FALSE}
secret_value <- 42
```

The chunk above ran silently. Proof: `secret_value` = `r secret_value`.

> Perfect for setup steps like loading packages.

## `message` and `warning`

```r
library(tidyverse)  # Normally prints startup messages
```

Set `message=FALSE` and `warning=FALSE` to keep output clean.

This is why the **setup chunk** at the top of every `.Rmd` includes:

```r
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Common Combinations

| Scenario | Options |
|----------|---------|
| Show code + output (default) | `echo=TRUE, eval=TRUE` |
| Results only (reports) | `echo=FALSE` |
| Show code, don't run | `eval=FALSE` |
| Run silently (setup) | `include=FALSE` |
| Clean figure | `echo=FALSE, warning=FALSE` |

## Exercise 2.1: Practice Options

Modify this chunk — change `echo=TRUE` to `echo=FALSE`, re-knit:

```{r exercise-echo, echo=TRUE}
chemicals <- c("PFOA", "PFOS", "PFNA", "PFHxS", "GenX")
half_life <- c(3.8, 5.4, 2.5, 8.5, NA)
data.frame(Chemical = chemicals, 
           `Half-life (yr)` = half_life, check.names = FALSE)
```

# Global Options {.segue}

## Set Defaults for the Whole Document

Instead of setting options on every chunk, use a **setup chunk**:

````
```{r setup, include=FALSE}`r ''`
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```
````

Every `.Rmd` should start with this. Override on individual chunks when needed.

# Inline Code {.segue}

## Embed R Results in Text

Syntax: `` `r knitr::inline_expr('expression')` ``

```
The mean was `r knitr::inline_expr('mean(c(2.3, 4.1, 3.7))')` µg/L.
```

Renders as: The mean was `r mean(c(2.3, 4.1, 3.7))` µg/L.

## Why Inline Code Matters

**Without inline code:**

> "We observed 247 participants with a mean age of 43.2 years."

Change the data → find and manually update every number. 😱

**With inline code:**

> "We observed `` `r knitr::inline_expr('n')` `` participants with a mean age of `` `r knitr::inline_expr('round(mean(ages), 1)')` `` years."

Change the data → re-knit → **everything updates automatically**. ✅

## Realistic Example

```{r inline-data}
set.seed(42)
n <- 150
ages <- round(rnorm(n, mean = 45, sd = 12))
pfoa <- round(rlnorm(n, meanlog = 1.2, sdlog = 0.8), 2)
```

> Our study enrolled `r n` participants with a mean age of
> `r round(mean(ages), 1)` years (SD = `r round(sd(ages), 1)`).
> Median serum PFOA was `r round(median(pfoa), 2)` ng/mL
> (IQR: `r round(quantile(pfoa, 0.25), 2)`–`r round(quantile(pfoa, 0.75), 2)`).

**Every number came from R code.**

## Formatting Inline Numbers

```{r}
big_number <- 1234567.89
```

Without formatting: `r big_number`

With `formatC()`: `r formatC(big_number, format = "f", big.mark = ",", digits = 0)`

```r
formatC(big_number, format = "f", big.mark = ",", digits = 0)
```

## Exercise 2.2: Dynamic Paragraph

Using this data, write a paragraph with inline code:

```{r exercise-inline}
set.seed(123)
bpa <- round(rlnorm(80, meanlog = 0.5, sdlog = 0.9), 2)
n_total <- length(bpa)
mean_bpa <- round(mean(bpa), 2)
pct_above <- round(mean(bpa > 5.0) * 100, 1)
```

Write: *"We measured BPA in `` `r knitr::inline_expr('n_total')` `` participants.
Mean concentration was `` `r knitr::inline_expr('mean_bpa')` `` µg/L.
`` `r knitr::inline_expr('pct_above')` ``% exceeded the reference dose."*

**Time: 5 minutes**
