---
title: "Module 4: Visualizations & Figure Control"
subtitle: "Publication-Quality Figures in R Markdown"
author: "NIEHS Training"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    css: rmd_slides_style.css
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "", fig.height = 2.8, fig.width = 7,
                      fig.align = "center", dpi = 150)
library(tidyverse)
```

```{r data, include=FALSE}
set.seed(42)
n <- 200
bio <- data.frame(
  age = round(rnorm(n, 48, 14)),
  sex = sample(c("Female", "Male"), n, replace = TRUE, prob = c(.55,.45)),
  pfoa = round(rlnorm(n, 0.8, 0.9), 2),
  cholesterol = round(rnorm(n, 210, 38)),
  site = sample(c("RTP", "Durham", "Chapel Hill"), n, replace = TRUE))
```

# Figures Are Code {.segue}

## The R Markdown Advantage

No more:

1. Make a plot in R
2. Save as PNG
3. Paste into Word
4. Data changes → repeat steps 1–3 for every figure 😩

With R Markdown: **the plot IS the code**. Re-knit and everything updates.

## Basic ggplot

```{r basic-plot, fig.height=2.8}
ggplot(bio, aes(x = pfoa, y = cholesterol)) +
  geom_point(alpha = 0.5) +
  labs(title = "PFOA vs. Cholesterol", 
       x = "PFOA (ng/mL)", y = "Cholesterol (mg/dL)")
```

## Publication-Quality Plot

```{r pub-plot, fig.height=2.6}
ggplot(bio, aes(x = pfoa, y = cholesterol, color = sex)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  scale_x_log10() +
  scale_color_manual(values = c("Female" = "#3182CE", "Male" = "#805AD5")) +
  labs(title = "Serum PFOA vs. Total Cholesterol",
       x = "Serum PFOA (ng/mL, log)", y = "Cholesterol (mg/dL)",
       color = "Sex") +
  theme_minimal(base_size = 11) + theme(legend.position = "bottom")
```

# Figure Chunk Options {.segue}

## Controlling Size and Appearance

| Option | Purpose | Example |
|--------|---------|---------|
| `fig.width` | Width in inches | `fig.width=8` |
| `fig.height` | Height in inches | `fig.height=3` |
| `fig.align` | Alignment | `"center"` |
| `dpi` | Resolution | `150` report, `300` publication |
| `fig.cap` | Caption text | `"Figure 1. My caption"` |
| `out.width` | Display width | `"80%"`, `"500px"` |

> `fig.width/height` = creation size. `out.width` = display size.

## Narrow and Tall

```{r narrow, fig.width=3.5, fig.height=3}
ggplot(bio, aes(x = sex, y = pfoa, fill = sex)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#3182CE", "#805AD5")) +
  scale_y_log10() + labs(y = "PFOA (ng/mL)", x = NULL) +
  theme_minimal() + theme(legend.position = "none")
```

## Wide and Short

```{r wide, fig.width=10, fig.height=2.5}
ggplot(bio, aes(x = age, y = cholesterol)) +
  geom_point(alpha = 0.3, color = "#4A5568") +
  geom_smooth(method = "loess", color = "#3182CE") +
  facet_wrap(~site, nrow = 1) +
  labs(x = "Age", y = "Cholesterol (mg/dL)") + theme_minimal(base_size = 9)
```

## Figure Captions

Use the `fig.cap` chunk option:

````
```{r, fig.cap="**Figure 1.** PFOA by site."}`r ''`
ggplot(bio, aes(x = site, y = pfoa)) + geom_boxplot()
```
````

```{r captioned, fig.cap="**Figure 1.** PFOA concentration by monitoring site.", echo=FALSE, fig.height=2.5}
ggplot(bio, aes(x = site, y = pfoa, fill = site)) +
  geom_boxplot(alpha = 0.7) + scale_fill_brewer(palette = "Set2") +
  scale_y_log10() + labs(x = NULL, y = "PFOA (ng/mL, log)") +
  theme_minimal() + theme(legend.position = "none")
```

## Multi-Panel with `patchwork`

```{r patchwork, fig.width=9, fig.height=2.5}
library(patchwork)
p1 <- ggplot(bio, aes(x = pfoa)) +
  geom_histogram(fill = "#3182CE", color = "white", bins = 25) +
  scale_x_log10() + labs(title = "A", x = "PFOA", y = "Count") + theme_minimal(base_size = 9)
p2 <- ggplot(bio, aes(x = pfoa, y = cholesterol)) +
  geom_point(alpha = 0.3) + geom_smooth(method = "lm", color = "#E53E3E") +
  scale_x_log10() + labs(title = "B", x = "PFOA", y = "Cholesterol") + theme_minimal(base_size = 9)
p1 + p2
```

## Interactive Plots with `plotly`

```{r plotly, fig.height=2.5}
library(plotly)
p <- ggplot(bio, aes(x = pfoa, y = cholesterol, color = sex,
                      text = paste("Age:", age))) +
  geom_point(alpha = 0.5) + scale_x_log10() +
  scale_color_manual(values = c("#3182CE","#805AD5")) +
  theme_minimal(base_size = 10)
ggplotly(p, tooltip = c("text", "x", "y"), height = 280)
```

> **Hover** over points! Only works in HTML output.

## Saving Figures for Manuscripts

```r
ggsave("output/figures/fig1_pfoa_cholesterol.png",
       width = 7, height = 5, dpi = 300)

# For journals requiring TIFF:
ggsave("output/figures/figure1.tiff",
       width = 7, height = 5, dpi = 300, compression = "lzw")
```

## Exercise 4.1: Multi-Panel Figure

Create two panels side by side:

- **A:** Histogram of `cholesterol`
- **B:** Scatterplot of `age` vs `cholesterol` colored by `sex`

Use `patchwork`. Set `fig.width=10, fig.height=3`. Add `fig.cap`.

**Time: 5 minutes**
