---
title: "Module 5: Scientific Writing in R Markdown"
subtitle: "Equations, Parameters, Citations, and Tabs"
author: "NIEHS Training"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    css: rmd_slides_style.css
    self_contained: true
params:
  chemical: "PFOA"
  threshold: 4.0
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "", fig.height = 2.8, fig.width = 7,
                      fig.align = "center")
library(tidyverse)
```

# LaTeX Equations {.segue}

## Inline Equations

Surround with single `$`:

| Syntax | Renders as |
|--------|-----------|
| `$\alpha = 0.05$` | $\alpha = 0.05$ |
| `$R^2 = 0.87$` | $R^2 = 0.87$ |
| `$\beta_1 = 0.42$` | $\beta_1 = 0.42$ |
| `$p < 0.001$` | $p < 0.001$ |
| `$\mu \pm \sigma$` | $\mu \pm \sigma$ |
| `$\text{BMI} = \frac{kg}{m^2}$` | $\text{BMI} = \frac{kg}{m^2}$ |

## Display Equations

Surround with double `$$` for centered, large equations:

```
$$Y_i = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \epsilon_i$$
```

$$Y_i = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \epsilon_i$$

## Equations for Environmental Health

**Hazard Quotient:**

$$\text{HQ} = \frac{\text{CDI}}{\text{RfD}}$$

**Lifetime Average Daily Dose:**

$$\text{LADD} = \frac{C \times IR \times EF \times ED}{BW \times AT}$$

**Hill Dose-Response Model:**

$$\text{Response} = \frac{E_{\max} \cdot C^n}{EC_{50}^n + C^n}$$

# Parameterized Reports {.segue}

## One Template, Many Reports

Define parameters in the YAML:

```yaml
params:
  chemical: "PFOA"
  threshold: 4.0
  study_year: 2024
```

Use anywhere: `params$chemical`, `params$threshold`

## Live Demo

This report analyzes **`r params$chemical`** with a threshold of
**`r params$threshold` ng/mL**.

```{r param-data, echo=FALSE, fig.height=2.3}
set.seed(42)
conc <- round(rlnorm(100, 0.8, 0.9), 2)
n_above <- sum(conc > params$threshold)
pct <- round(n_above / length(conc) * 100, 1)
ggplot(data.frame(x = conc), aes(x = x)) +
  geom_histogram(fill = "#3182CE", color = "white", bins = 25) +
  geom_vline(xintercept = params$threshold, color = "#E53E3E",
             linewidth = 1, linetype = "dashed") +
  annotate("text", x = params$threshold + 0.5, y = Inf,
           label = paste0(pct, "% above threshold"),
           vjust = 2, hjust = 0, color = "#E53E3E", size = 3.5) +
  labs(title = paste(params$chemical, "Distribution"),
       x = paste(params$chemical, "(ng/mL)"), y = "Count") +
  theme_minimal()
```

`r pct`% of participants exceeded the `r params$threshold` ng/mL threshold.

## Re-Rendering with Different Parameters

```r
rmarkdown::render(
  "my_report.Rmd",
  params = list(chemical = "PFOS", threshold = 6.0),
  output_file = "report_PFOS.html"
)
```

**One `.Rmd` → 50 site reports, 10 chemical reports, etc.**

Also: **Knit button dropdown → "Knit with Parameters..."** for an interactive dialog.

# Tabbed Sections {.segue}

## Organize Content with Tabs

Add `{.tabset}` to a heading to turn sub-headings into tabs:

```markdown
## Results {.tabset}

### Table
(table content here)

### Figure
(figure content here)

### Summary
(summary content here)
```

Readers click tabs to switch between views. HTML output only.

# Citations {.segue}

## Proper Academic Citations

1. Create a **BibTeX file** (`references.bib`):

```bibtex
@article{smith2023pfas,
  title = {PFAS Exposure and Health Outcomes},
  author = {Smith, Jane and Doe, John},
  journal = {Environ Health Perspect},
  year = {2023},
  volume = {131}, pages = {1--12}
}
```

2. Point to it in the YAML:

```yaml
bibliography: references.bib
csl: apa.csl
```

## Citing in Text

| Syntax | Result |
|--------|--------|
| `@smith2023pfas` | Smith and Doe (2023) |
| `[@smith2023pfas]` | (Smith and Doe, 2023) |
| `[@smith2023; @jones2024]` | (Smith, 2023; Jones, 2024) |
| `[-@smith2023pfas]` | (2023) — suppress author |

A **References** section is auto-generated at the document end.

Download CSL files for specific journals from [zotero.org/styles](https://www.zotero.org/styles).

## Exercise 5.1

1. Write the odds ratio equation: $OR = \frac{a/b}{c/d}$ using `$$`
2. Change `chemical` to `"PFOS"` and `threshold` to `6.0` in the YAML
3. Re-knit and see everything update

**Time: 5 minutes**
